Quantifying Heterogeneity in SAR Imagery with the Rényi Entropy

### 2025-06-03 Text/Articles/RENYI_ENTROPY_SAR_2025/Reviews/Responses-Reviewer1.tex

We thank the reviewer for raising this important point. While non-parametric entropy estimation is indeed present in the literature, we have not found prior works that apply a non-parametric estimator of the Rényi entropy specifically for heterogeneity detection in SAR imagery.

Our contribution lies in formulating a hypothesis test that uses the entropy estimator as a test statistic to distinguish between homogeneous and heterogeneous regions. Classical parametric inference alternatives, such as likelihood ratio or Wald tests, are not applicable here due to the limiting behavior of the roughness parameter (\(\alpha \to -\infty\)) under the null hypothesis.

In addition, we enhance the non-parametric Rényi entropy estimator by incorporating a bootstrap based bias correction technique. This resampling strategy not only reduces the estimator’s bias but also improves the accuracy of entropy estimation, particularly when working with small sample sizes. This improvement is especially relevant in practical SAR image analysis, where estimations are commonly performed over small windows, such as \(7 \times 7\) pixels.

Moreover, our aim is to propose a statistically grounded, interpretable approach that provides analytical significance measures. This distinguishes our method from data-driven alternatives, and makes it especially useful for unsupervised analysis or scenarios with limited ground truth information.

@@

\end{responsebox}

\vspace{2em}
\begin{reviewbox}{Comment 5}
5. Fig 6, 7 and 8 show the detection of heterogeneous areas. However, a statistical measure indicating the detection performance is absent.
\end{reviewbox}

\begin{responsebox}{Response}
In the revised manuscript, we have included a quantitative performance evaluation to complement the visual comparison.

Specifically, we manually defined eight polygonal regions of interest (ROIs) in each SAR image: four clearly homogeneous areas (e.g., water, flat croplands) and four clearly heterogeneous areas (e.g., urban centers). Each ROI was labeled as either homogeneous (class 0, shown in blue) or heterogeneous (class 1, shown in red), as illustrated in Figs.~5a, 6a, and 7a (corresponding to Figs.~6a, 7a, and 8a in the original submission).

These ROIs were then rasterized to create a partial ground truth mask. The $p$-value maps were thresholded at a 0.05 significance level to generate binary decision maps: values below this threshold were labeled as heterogeneous (1), and those above or equal to it as homogeneous (0). The resulting prediction maps were compared against the ground truth labels within the ROIs.

Based on this comparison, we computed standard performance metrics, including F1-score, kappa coefficient, and overall accuracy, summarized in Table~III.

Additionally, to evaluate performance across all decision thresholds, we computed ROC curves and the corresponding AUC values for each method, as shown in Fig.~9.

\begin{quote}
\textcolor{blue}{These additions appear in Section~IV-C of the revised manuscript.}
\end{quote}
\end{responsebox}

### 2025-06-03 Text/Articles/RENYI_ENTROPY_SAR_2025/R1-Renyi-entropy-sar.qmd
R1 submission: clean and tracked versions, updated responses

```
\renewcommand{\tablename}{TABLE}
# Introduction
[S]{.IEEEPARstart}[ynthetic]{}
 Aperture Radar (SAR) technology has become essential for many applications&nbsp;[@Mondini2021]. 
 It provides high-resolution data independent of sunlight and operates in various weather conditions that facilitate global Earth monitoring&nbsp;[@Zeng2020].
However, the effective use of SAR data depends on understanding its statistical properties because it is affected by speckle, a noise-like interference effect&nbsp;[@Baraha2023]. In intensity format, speckle is non-Gaussian, which  complicates image analysis.

@@

Recent studies highlight the role of entropy in SAR classification. 
Cassetti et al. [@Cassetti2022] evaluated entropy estimators in both supervised and unsupervised models. 
Gallet et al. [@Gallet2024] proposed a Rényi divergence-based framework for explainable classification. 
Parikh et al. [@Parikh2019] discussed challenges in deep learning for SAR, including limited availability of labeled data and the complexity of tuning hyperparameters. These works motivate lightweight, interpretable alternatives.

We propose a statistical test based on a non-parametric estimator of Rényi entropy to identify heterogeneous regions in SAR data.
The test assesses whether the observed Rényi entropy significantly differs from its expected theoretical value under the homogeneity assumption, improving upon our previous Shannon-based approach&nbsp;[@Frery2024].
\textcolor{blue}{We further enhance the non-parametric estimator through a bootstrap-based bias correction, improving accuracy for small windows.}
\textcolor{blue}{The proposed approach is unsupervised, requires no training data, and yields interpretable $p$-values, making it especially suitable for scenarios where ground-truth information is limited or unavailable.}

The rest of the article is organized as follows.
Section&nbsp;\ref{sec:pre} overviews the statistical models for SAR intensity data and Rényi entropy.
Section&nbsp;\ref{sec:met} describes the proposed hypothesis test.
Section&nbsp;\ref{sec:app} evaluates the performance of the test using SAR data.
Finally, the conclusions are presented in Section&nbsp;\ref{sec:conclusion}.

@@

\textcolor{blue}{where the excess term $\Delta_\alpha$ depends on the roughness parameter.
When $\alpha\!\to\!-\infty$ (fully developed speckle), this excess tends to zero,
%<!-- , so the heterogeneous model collapses to the homogeneous one in Equation \eqref{eq-HGammaSAR}.  -->
and detecting heterogeneity is equivalent to deciding whether $\Delta_\alpha$ is significantly different from zero.}
<!-- Note that the Rényi entropy of $\mathcal{G}^0_I$ can be expressed in terms of that of $\Gamma_{\mathrm{SAR}}$, with additional terms that depend on the roughness parameter $\alpha$. As $\alpha \to -\infty$, these terms tend to zero, reducing&nbsp;\eqref{eq-HGI0} to&nbsp;\eqref{eq-HGammaSAR}. -->
<!-- This work's core idea is to check if this excess entropy is significantly different from zero. -->
Fig.&nbsp;\ref{fig-convergence} illustrates this behaviour: as $\alpha$ decreases the coloured curves of $H_\lambda(\mathcal{G}^0_I)$ converge towards the solid black curve of $H_\lambda(\Gamma_{\mathrm{SAR}})$.
<!-- , confirming the limiting behavior. -->

@@

\textcolor{blue}{
We aim to determine the optimal order $\lambda$ for the Rényi entropy estimator using simulated samples of size $n = 49$, drawn from  $Z\sim \Gamma_{\text{SAR}}(5,1)$ distribution.
We computed the bias and the mean squared error (MSE) for vaying $\lambda$ with a Monte Carlo experiment.
We found that for $L > 1$, $\lambda = 0.9$  yields the lowest MSE while maintaining a low bias, thus offering a favorable trade-off between bias and variance. 
As shown in Fig. \ref{fig-optimal_order}, $\lambda = 0.85$ produces slightly lower bias, but leads to higher MSE, supporting the choice of $\lambda = 0.9$.}

@@

```{=latex}
\begin{figure}[hbt]
  \centering
  \includegraphics[width=0.42\textwidth]{Figures-R1/fig-optimal_order-1.pdf}
  \vspace{0.5em}
  \caption{Bias and MSE as a function of $\lambda$, with $n=49$, $L=5$.}
  \label{fig-optimal_order}
\end{figure}
```
Particularly, for $L = 1$, the optimal $\lambda$ tends to be higher, and we chose $\lambda = 3$ to achieve good results.

@@

Following Refs.&nbsp;[@Frery2024;@Alpala2024], we refine the non-parametric entropy estimator $\widehat{H}_{\lambda}$ in&nbsp;\eqref{eq:est_R} with bootstrap, obtaining $\widetilde{H}_{\lambda}$:
\begin{equation*}
\widetilde{H}_{\lambda} = 2\widehat{H}_{\lambda}(\bm{Z}) - \frac{1}{B} \sum_{b=1}^{B} \widehat{H}_{\lambda}(\bm{Z}^{(b)}),
\end{equation*}
where $B$ is the number of bootstrap replications, and $\bm{Z}^{(b)}$ is the $b$-th resampled dataset obtained by drawing $n$ observations with replacement from $\bm{Z}$.

A Monte Carlo study with $1000$ replications for each sample size $n \in \{9, 25, 49, 81, 121\}$ from the $\Gamma_{\text{SAR}}$ (5,1) confirms that for $\lambda=0.9$, the bootstrap-corrected estimator $\widetilde{H}_{\lambda}$ ($B=200$) reduces both bias and MSE compared to the original $\widehat{H}_{\lambda}$, with significant improvements for small sample sizes, as shown in Fig. \ref{fig-bias_mse}.
```{r Simulated_data_bias_B1, echo=FALSE, message=FALSE, cache = TRUE, autodep = TRUE}
set.seed(1234567890, kind = "Mersenne-Twister")

@@

## The Proposed Test
The test statistic quantifies the discrepancy between estimated and theoretical entropy under homogeneity. Since $L\geq1$ is known, we define the test statistic as follows:
\begin{multline}
\label{eq-test}
S_{\widetilde{H}_{\lambda}}(\bm{Z}; L) = \widetilde{H}_{\lambda} - \bigl\{\ln \widehat{\mu} - \ln L + \frac{1}{1-\lambda}
\bigl[-\lambda\,\ln\Gamma(L) \\ 
+ \ln\Gamma\bigl(\lambda(L-1)+1\bigr)  
- \bigl(\lambda(L-1)+1\bigr)\,\ln\lambda
\bigr]\bigr\},
\end{multline}
where $\widehat{\mu}={n}^{-1}\sum_{i=1}^n Z_{i}$ is the sample mean.
The test statistic can be interpreted as:
$$
S_{\widetilde H_\lambda} 
= 
\underbrace{\widetilde H_\lambda}_{\text{estimated}} 
\;-\;
\underbrace{H_\lambda\bigl(\Gamma_{\mathrm{SAR}}\bigr)}_{\text{theoretical under } H_0}\hspace{-0.5em},
$$
\textcolor{blue}{the difference between the estimated entropy and the expected value under homogeneity. Values close to zero indicate that the region behaves like fully developed speckle, while large positive values signal excess entropy and, thus, heterogeneity.
This formulation avoids estimating  $\mathcal{G}^0_I$ parameters  such as $\alpha$, offering a simple, interpretable, and statistically grounded test. Moreover, the method remains effective for small samples, aided by the bootstrap bias correction.}

@@

<!-- ```{=latex} -->
<!-- \begin{figure}[hbt] -->
<!--   \centering -->
<!--   \includegraphics[width=0.35\textwidth]{Figures-R1/fig-density_standardized-1.pdf} -->
<!--   %\vspace{0.1em} -->
<!--   \caption{Standardized empirical densities of $S_{\widetilde{H}_{\lambda}}(\bm{Z}; L)$ under $\mathcal{H}_0$.} -->
<!--   \label{fig-density_standardized} -->
<!-- \end{figure} -->
<!-- ``` -->
In general, hypotheses tests aim to control the Type&nbsp;I error rate (size) with high test power (sensitivity to departures from $\mathcal{H}_0$, low Type&nbsp;II error rate).
To assess these properties, we performed a Monte Carlo simulation with $1000$ replications at significance levels \SI{1}{\percent}, \SI{5}{\percent}, and \SI{10}{\percent}, evaluating the test under the null hypothesis ($\Gamma_{\text{SAR}}$ distribution), varying sample size and values of $L$ for $\mu=1$. 
The observed Type&nbsp;I error rates align with the nominal values, confirming the test validity.

@@

print(table_combined_result)
```
\textcolor{blue}{
We applied the proposed test to SAR images using sliding windows of $7 \times 7$ pixels. 
For each window, we computed the test statistic and calculated the $p$-value that quantifies the statistical evidence against local homogeneity. 
High $p$-values, e.g., $p > 0.05$, indicate no evidence to reject homogeneity, while low values, e.g., $p < 0.05$, suggest statistically significant deviations and are interpreted as heterogeneity.
}
<!-- Note that the visual outputs in Figs.~\ref{fig:london}--\ref{fig:dublin} show the $p$-values obtained from the hypothesis testing procedure, not the test statistic or entropy values. -->

@@

SAR_data <- data.frame(
  Site = c("London", "Munich", "Dublin" ),
  Mission = c("TanDEM-X", "UAVSAR", "TanDEM-X"),
  Band = c("X", "L", "X"),
  Size = c("$2000\\times2000$", "$1024\\times1024$", "$1100\\times1100$"),
  L = c(1, 12, 16),
  Resol = c("$0.99/0.99$", "$4.9/7.2$", "$1.35/1.35$"),
  Date = c("12-11-2021", "16-04-2015", "03-09-2017")
)

@@

```
\textcolor{blue}{
For each SAR image, the number of looks $L$ was obtained from the product of azimuth and range looks provided in the metadata (SNAP). 
We validated this nominal value with the equivalent number of looks (ENL) from manually selected homogeneous areas using the classical formula $\text{ENL} = 1/\widehat{\text{CV}}^2$, based on the sample coefficient of variation. 
The results were consistent with the metadata. 
We also verified that moderate deviations in $L$ did not significantly impact the outcome of the test, confirming the robustness of the proposed method.}

@@

<!-- New color  Palettes viridis B (inferno) -->
```{=latex}
\begin{figure*}[hbt]
    \centering
    \begin{subfigure}{0.145\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/london_roi.png}
        \caption{SAR image}
        \label{fig:london-sar}
    \end{subfigure}
    \hspace{0.00001\textwidth}
    \begin{subfigure}{0.142\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/london_optical.png}
        \caption{Optical image}
        \label{fig:london-optical}
    \end{subfigure}
    \hspace{0.00001\textwidth}
    \begin{subfigure}{0.179\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/p-values_renyi-london_B1.png}
        \caption{$p$-value map for $\small{S_{\widetilde{H}_{\lambda}}}$}
        \label{fig:london-renyi}
    \end{subfigure}
    \hspace{0.00001\textwidth}
    \begin{subfigure}{0.145\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/H_005_london_renyi_L1_.png}
        \caption{Binary map for $\small{S_{\widetilde{H}_{\lambda}}}$}
        \label{fig:london-005-renyi}
    \end{subfigure}
   \hspace{0.00001\textwidth}
    \begin{subfigure}{0.18\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/p-values_shannon-london_B1.png}
        \caption{$p$-value map for $\tiny{S_{\widetilde{H}_{\text{AO}}}}$ }
        \label{fig:london-shann}
    \end{subfigure}
    \hspace{0.00001\textwidth}
    \begin{subfigure}{0.145\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/H_005_london_shannon.png}
        \caption{Binary map for $\small{S_{\widetilde{H}_{\text{AO}}}}$}
        \label{fig:london-005-shann}
    \end{subfigure}
    \caption{Detection of heterogeneous areas in a SAR image over London, UK: comparison with the tests $\small{S_{\widetilde{H}_{\lambda}}}$ and $\small{S_{\widetilde{H}_{\text{AO}}}}$.}
    \label{fig:london}
\end{figure*}
```

```{=latex}
\begin{figure*}[hbt]
    \centering
        \begin{subfigure}{0.145\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/munich_roi.png}
        \caption{SAR image}
        \label{fig:munich-sar}
    \end{subfigure}
    \hspace{0.00001\textwidth}
  \begin{subfigure}{0.137\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/munich_optical.png}
        \caption{Optical image}
        \label{fig:munich-optical}
    \end{subfigure}
    \hspace{0.00001\textwidth}
    \begin{subfigure}{0.178\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/p-values_renyi-munich-B.png}
        \caption{$p$-value map for $\small{S_{\widetilde{H}_{\lambda}}}$}
        \label{fig:munich-renyi}
    \end{subfigure}
    \hspace{0.00001\textwidth}
    \begin{subfigure}{0.144\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/H_005_renyi_munich.png}
        \caption{Binary map for $\small{S_{\widetilde{H}_{\lambda}}}$}
        \label{fig:munich-005-renyi}
    \end{subfigure}
    \hspace{0.00001\textwidth}
    \begin{subfigure}{0.178\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/p-values_shannon-munich-B.png}
        \caption{$p$-value map for $S_{\widetilde{H}_{\text{AO}}}$}
        \label{fig:munich-shann}
    \end{subfigure}
   \hspace{0.00001\textwidth}
    \begin{subfigure}{0.144\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/H_005_shannon_munich.png}
        \caption{Binary map for $\small{S_{\widetilde{H}_{\text{AO}}}}$}
        \label{fig:munich-005-shann}
    \end{subfigure}
    \caption{Detection of heterogeneous areas in a SAR image over Munich, Germany: comparison with the tests $\small{S_{\widetilde{H}_{\lambda}}}$ and $\small{S_{\widetilde{H}_{\text{AO}}}}$.}
    \label{fig:munich}
\end{figure*}

```{=latex}
\begin{figure*}[hbt]
    \centering
        \begin{subfigure}{0.145\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/dublin_roi3.png}
        \caption{SAR image}
        \label{fig:dublin-sar}
    \end{subfigure}
    \hspace{0.00001\textwidth}
\begin{subfigure}{0.145\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/dublin.png}
        \caption{Optical image}
        \label{fig:dublin-optical}
    \end{subfigure}
    \hspace{0.00001\textwidth}
    \begin{subfigure}{0.178\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/p-values_renyi-dublin_B.png}
        \caption{$p$-value map $\small{S_{\widetilde{H}_{\lambda}}}$}
        \label{fig:dublin-renyi}
    \end{subfigure}
    \hspace{0.00001\textwidth}
    \begin{subfigure}{0.144\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/H_005_dublin_renyi.png}
        \caption{Binary map $\small{S_{\widetilde{H}_{\lambda}}}$}
        \label{fig:dublin-005-renyi}
    \end{subfigure}
    \hspace{0.00001\textwidth}
    \begin{subfigure}{0.178\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/p-values_shannon-dublin_B.png}
        \caption{$p$-value $\small{S_{\widetilde{H}_{\text{AO}}}}$}
        \label{fig:dublin-shann}
    \end{subfigure}
    \hspace{0.0000001\textwidth}
    \begin{subfigure}{0.144\textwidth}
        \includegraphics[width=\linewidth]{./Figures-R1/H_005_dublin-shannon.png}
        \caption{Binary $\small{S_{\widetilde{H}_{\text{AO}}}}$}
        \label{fig:dublin-005-shann}
    \end{subfigure}
    \caption{Detection of heterogeneous areas in a SAR image over Dublin Port, Ireland: comparison with the tests $\small{S_{\widetilde{H}_{\lambda}}}$ and $\small{S_{\widetilde{H}_{\text{AO}}}}$.}
    \label{fig:dublin}
\end{figure*}

@@

Eight polygonal regions of interest (ROIs) were manually selected for each scene (Figs.&nbsp;\ref{fig:london-sar}-\ref{fig:dublin-sar}): heterogeneous areas (red, class 1) and homogeneous areas (blue, class 0). These ROIs were rasterized to match the SAR resolution, forming a partial ground-truth. 
We thresholded the $p$-value maps at 0.05, labeling each pixel as heterogeneous (1) or homogeneous (0) and then compared these decision maps with the ground-truth to compute: F1-score ($F_1$), Kappa coefficient ($\kappa$), and overall accuracy (OA). Table&nbsp;\ref{tab:table_metrics} summarizes the results.
Across all scenes, the Rényi-based test outperformed Shannon’s in all metrics, with notable gains in London ($F_1$: 0.695 vs. 0.617; $\kappa$: 0.626 vs. 0.542), highlighting its ability to capture  heterogeneity under strong speckle (single-look).

@@

\textcolor{blue}{Using SAR data, we compared the Rényi-based test with a Shannon-based counterpart through $p$-value maps and binary decisions. Visual analysis showed greater sensitivity of the Rényi-based test to textural variations.
Quantitative evaluation with ground-truth ROIs and standard classification metrics confirmed the superior performance of the Rényi-based test across all scenes.}

\textcolor{blue}{Although machine learning-based SAR classifiers are widely used, our focus is on interpretable, statistically grounded methods that yield analytical significance measures. This makes them suitable for rapid detection, screening, and environments with limited data availability.}

\textcolor{blue}{Future work will explore the integration of the improved Rényi estimator into SAR classification pipelines, both supervised and unsupervised, extending its utility beyond heterogeneity testing.}

>>>>

```
\renewcommand{\tablename}{TABLE}
# Introduction
[S]{.IEEEPARstart}[ynthetic]{}
 Aperture Radar (SAR) technology has become essential for many applications&nbsp;[@Mondini2021]. 
 It provides high-resolution data independent of sunlight and operates in various weather conditions, facilitating global Earth monitoring&nbsp;[@Zeng2020].
However, the effective use of SAR data depends on understanding its statistical properties because it is affected by speckle, a noise-like interference effect&nbsp;[@Baraha2023]. In intensity format, speckle is non-Gaussian, which complicates image analysis.

@@

Recent studies highlight the role of entropy in SAR classification. 
Cassetti et al.&nbsp;[@Cassetti2022] evaluated entropy estimators in both supervised and unsupervised models. 
Gallet et al.&nbsp;[@Gallet2024] proposed a Rényi divergence-based framework for explainable classification. 
Parikh et al.&nbsp;[@Parikh2019] discussed challenges in deep learning for SAR, including limited availability of labeled data and the complexity of tuning hyperparameters. These works motivate lightweight, interpretable alternatives.

We propose a statistical test based on a non-parametric estimator of Rényi entropy to identify heterogeneous regions in SAR data.
The test assesses whether the observed Rényi entropy significantly differs from its expected theoretical value under the homogeneity assumption, improving upon our previous Shannon-based approach&nbsp;[@Frery2024].
We further enhance the non-parametric estimator through a bootstrap-based bias correction, improving accuracy for small windows.
The proposed approach is unsupervised, requires no training data, and yields interpretable $p$-values, making it especially suitable for scenarios where ground truth information is limited or unavailable.

The rest of this article is organized as follows.
Section&nbsp;\ref{sec:pre} overviews the statistical models for SAR intensity data and Rényi entropy.
Section&nbsp;\ref{sec:met} describes the proposed hypothesis test.
Section&nbsp;\ref{sec:app} evaluates the performance of the test using SAR data.
Finally, the conclusions are presented in Section&nbsp;\ref{sec:conclusion}.

@@

where the excess term $\Delta_\alpha$ depends on the roughness parameter.
When $\alpha\!\to\!-\infty$ (fully developed speckle), this excess tends to zero,
%<!-- , so the heterogeneous model collapses to the homogeneous one in Equation \eqref{eq-HGammaSAR}.  -->
and detecting heterogeneity is equivalent to deciding whether $\Delta_\alpha$ is significantly different from zero.
<!-- Note that the Rényi entropy of $\mathcal{G}^0_I$ can be expressed in terms of that of $\Gamma_{\mathrm{SAR}}$, with additional terms that depend on the roughness parameter $\alpha$. As $\alpha \to -\infty$, these terms tend to zero, reducing&nbsp;\eqref{eq-HGI0} to&nbsp;\eqref{eq-HGammaSAR}. -->
<!-- This work's core idea is to check if this excess entropy is significantly different from zero. -->
Fig.&nbsp;\ref{fig-convergence} illustrates this behaviour: as $\alpha$ decreases the colored curves of $H_\lambda(\mathcal{G}^0_I)$ converge towards the solid black curve of $H_\lambda(\Gamma_{\mathrm{SAR}})$.
<!-- , confirming the limiting behavior. -->

@@

We aim to determine the optimal order $\lambda$ for the Rényi entropy estimator using simulated samples of size $n = 49$ from  $Z\sim \Gamma_{\text{SAR}}(5,1)$.
We computed the bias and the mean squared error (MSE) for varying $\lambda$ via a Monte Carlo experiment.
We found that for $L > 1$, $\lambda = 0.9$  yields the lowest MSE while maintaining a low bias, thus offering a favorable trade-off between bias and variance. 
As shown in Fig.&nbsp;\ref{fig-optimal_order},&nbsp;$\lambda = 0.85$ produces slightly lower bias, but results in higher MSE, supporting the choice of $\lambda = 0.9$.

@@

```{=latex}
\begin{figure}[hbt]
  \centering
  \includegraphics[width=0.42\textwidth]{Figures-R1/fig-optimal_order-1.pdf}
  \vspace{0.5em}
  \caption{Bias and MSE as a function of $\lambda$, with $n=49$, $L=5$.}
  \label{fig-optimal_order}
\end{figure}
```
Particularly, for $L = 1$, the optimal $\lambda$ tends to be higher, and we choose $\lambda = 3$ to achieve good results.

@@

Following Refs.&nbsp;[@Frery2024;@Alpala2024], we refine the non-parametric entropy estimator $\widehat{H}_{\lambda}$ in&nbsp;\eqref{eq:est_R} with bootstrap, obtaining $\widetilde{H}_{\lambda}$:
\begin{equation*}
\widetilde{H}_{\lambda} = 2\widehat{H}_{\lambda}(\bm{Z}) - \frac{1}{B} \sum_{b=1}^{B} \widehat{H}_{\lambda}(\bm{Z}^{(b)}),
\end{equation*}
where $B$ is the number of bootstrap replications, and $\bm{Z}^{(b)}$ denotes  the $b$-th resampled dataset obtained by drawing $n$ observations with replacement from $\bm{Z}$.

A Monte Carlo study with $1000$ replications for each sample size $n \in \{9, 25, 49, 81, 121\}$ from the $\Gamma_{\text{SAR}}$ (5,1) confirms that for $\lambda=0.9$, the bootstrap-corrected estimator $\widetilde{H}_{\lambda}$ ($B=200$) reduces both bias and MSE compared to the original $\widehat{H}_{\lambda}$, with significant improvements for small sample sizes, as shown in Fig.&nbsp;\ref{fig-bias_mse}.
```{r Simulated_data_bias_B1, echo=FALSE, message=FALSE, cache = TRUE, autodep = TRUE}
set.seed(1234567890, kind = "Mersenne-Twister")

@@

## The Proposed Test
The test statistic quantifies the discrepancy between estimated and theoretical entropy under homogeneity. Since $L\geq1$ is known, we define the test statistic as follows:
\begin{multline}
\label{eq-test}
S_{\widetilde{H}_{\lambda}}(\bm{Z}; L) = \widetilde{H}_{\lambda} - \bigl\{\ln \widehat{\mu} - \ln L + \frac{1}{1-\lambda}
\bigl[-\lambda\,\ln\Gamma(L) \\ 
+ \ln\Gamma\bigl(\lambda(L-1)+1\bigr)  
- \bigl(\lambda(L-1)+1\bigr)\,\ln\lambda
\bigr]\bigr\},
\end{multline}
where $\widehat{\mu}={n}^{-1}\sum_{i=1}^n Z_{i}$ is the sample mean.
The test statistic can be interpreted as:
$$
S_{\widetilde H_\lambda} 
= 
\underbrace{\widetilde H_\lambda}_{\text{estimated}} 
\;-\;
\underbrace{H_\lambda\bigl(\Gamma_{\mathrm{SAR}}\bigr)}_{\text{theoretical under } H_0}\hspace{-0.5em},
$$
the difference between the estimated entropy and the expected value under homogeneity. Values close to zero indicate that the region behaves like fully developed speckle, while large positive values signal excess entropy and, thus, heterogeneity.
This formulation avoids the need to estimate $\mathcal{G}^0_I$ parameters  such as $\alpha$, offering a simple, interpretable, and statistically grounded test. Moreover, the method remains effective for small samples, aided by the bootstrap bias correction.

@@

<!-- ```{=latex} -->
<!-- \begin{figure}[hbt] -->
<!--   \centering -->
<!--   \includegraphics[width=0.35\textwidth]{Figures-R1/fig-density_standardized-1.pdf} -->
<!--   %\vspace{0.1em} -->
<!--   \caption{Standardized empirical densities of $S_{\widetilde{H}_{\lambda}}(\bm{Z}; L)$ under $\mathcal{H}_0$.} -->
<!--   \label{fig-density_standardized} -->
<!-- \end{figure} -->
<!-- ``` -->
In general, hypothesis tests aim to control the Type&nbsp;I error rate (size) with high test power (sensitivity to departures from $\mathcal{H}_0$, low Type&nbsp;II error rate).
To assess these properties, we performed a Monte Carlo simulation with $1000$ replications at significance levels \SI{1}{\percent}, \SI{5}{\percent}, and \SI{10}{\percent}, evaluating the test under the null hypothesis ($\Gamma_{\text{SAR}}$ distribution), varying sample size and values of $L$ for $\mu=1$. 
The observed Type&nbsp;I error rates align with the nominal values, confirming the test validity.

@@

print(table_combined_result)
```

We applied the proposed test to SAR images using sliding windows of $7 \times 7$ pixels. 
For each window, we computed the test statistic and calculated the $p$-value that quantifies the statistical evidence against local homogeneity. 
High $p$-values, e.g., $p > 0.05$, indicate no evidence to reject homogeneity, while low values, e.g., $p < 0.05$, suggest statistically significant deviations and are interpreted as heterogeneity.

<!-- Note that the visual outputs in Figs.~\ref{fig:london}--\ref{fig:dublin} show the $p$-values obtained from the hypothesis testing procedure, not the test statistic or entropy values. -->

@@

SAR_data <- data.frame(
  Site = c("London", "Munich", "Dublin" ),
  Mission = c("TanDEM-X", "UAVSAR", "TanDEM-X"),
  Band = c("X", "L", "X"),
  Size = c("$2000\\times2000$", "$1024\\times1024$", "$1100\\times1100$"),
  L = c(1, 12, 16),
  Resol = c("$0.99\\times0.99$", "$4.9\\times7.2$", "$1.35\\times1.35$"),
  Date = c("12-11-2021", "16-04-2015", "03-09-2017")
)

@@

For each SAR image, the number of looks $L$ was obtained from the product of azimuth and range looks provided in the metadata (SNAP). 
We validated this nominal value with the equivalent number of looks (ENL) from manually selected homogeneous areas using the classical formula $\text{ENL} = 1/\widehat{\text{CV}}^2$, based on the sample coefficient of variation. 
The results were consistent with the metadata. 
We also verified that moderate deviations in $L$ did not significantly impact the outcome of the test, confirming the robustness of the proposed method.

@@

@@

Eight polygonal regions of interest (ROIs) were manually selected for each scene (Figs.&nbsp;\ref{fig:london-sar}-\ref{fig:dublin-sar}): heterogeneous areas (red, class 1) and homogeneous areas (blue, class 0). These ROIs were rasterized to match the SAR resolution, forming a partial ground truth. 
We thresholded the $p$-value maps at 0.05, labeling each pixel as heterogeneous (1) or homogeneous (0) and then compared these decision maps with the ground truth to compute: F1-score ($F_1$), Kappa coefficient ($\kappa$), and overall accuracy (OA). Table&nbsp;\ref{tab:table_metrics} summarizes the results.
Across all scenes, the Rényi-based test outperformed Shannon’s in all metrics, with notable gains in London ($F_1$: 0.695 vs. 0.617; $\kappa$: 0.626 vs. 0.542), highlighting its ability to capture  heterogeneity under strong speckle (single-look).

@@

Using SAR data, we compared the Rényi-based test with a Shannon-based counterpart through $p$-value maps and binary decisions. Visual analysis showed greater sensitivity of the Rényi-based test to textural variations.
Quantitative evaluation with ground truth ROIs and standard classification metrics confirmed the superior performance of the Rényi-based test across all scenes.

Although machine learning-based SAR classifiers are widely used, our focus is on interpretable, statistically grounded methods that yield analytical significance measures. This makes them suitable for rapid detection, screening, and environments with limited data availability.

Future work will explore the integration of the improved Rényi estimator into SAR classification pipelines, both supervised and unsupervised, extending its utility beyond heterogeneity testing.
